---
title: "final ML"
author: "Zac Macintyre"
date: "6/9/2021"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



```{r}
library(boot)
library(mltools)
library(data.table)
library(psych)
library(glmnet)
fires = read.csv("ForestFires.csv")
head(fires)
```

```{r}

colSums(is.na(fires))

```
As we said in the presentation there is no missing values just doing a little check

# This section is me doing some initial data visualization and some data analysis

```{r}
barplot(table(fires$month))
barplot(table(fires$day))
```
```{r}
summary(fires)
```

```{r}
par(mfrow=c(2,2))
plot(fires$rain, fires$area)
plot(fires$temp, fires$area)
plot(fires$FFMC, fires$area)
plot(fires$DMC, fires$area)
```
```{r}
par(mfrow=c(2,2))
plot(fires$DC, fires$area)
plot(fires$ISI, fires$area)
plot(fires$wind, fires$area)
plot(fires$RH, fires$area)
```

Some summary statistics
```{r}
summary(fires)
describe(fires)
```

Correlation
```{r}
fires_matirx = fires[,-c(3,4)]
corelation_m = cor(fires_matirx, use='pairwise.complete.obs')
lower = lower.tri(fires_matirx)
corelation_m
hist(corelation_m[lower], xlab = 'correlations of lower matrix')
```

# In this section I will start actually looking into some regression settings

As suggested in Cortez and Morais (2007), transforming area to log(area+1) can correct area's skewness toward zero

```{r}
log_fires = fires
log_fires$log_area = log(log_fires$area +1)
log_fires = log_fires[, -c(13)]
```


This is for fires not using X,Y,Day,Month
```{r}
fires_lm = log_fires[,-c(1,2,3,4)]
#head(log_fires_lm)
#fitting the most generic model
set.seed(1)
test = sample(nrow(fires_lm), floor(nrow(fires_lm)*.2))
linear_fit = lm(log_area~., data = fires_lm[-test,])
summary(linear_fit)

# getting MSE for the test
pred = predict(linear_fit, fires_lm[test,])
mean((pred - fires_lm[test, "log_area"])**2)

```

This is for fires not using Day,Month
```{r}
fires_lm = log_fires[,-c(3,4)]
#head(fires_lm)
#fitting the most generic model
test = sample(nrow(fires_lm), floor(nrow(fires_lm)*.2))
linear_fit = lm(log_area~., data = fires_lm[-test,])
summary(linear_fit)


# getting MSE for the test
pred = predict(linear_fit, fires_lm[test,])
mean((pred - fires_lm[test, "log_area"])**2)
```


```{r}
fires$month = as.factor(log_fires$month)
fires$day = as.factor(log_fires$day)
newdata = one_hot(as.data.table(log_fires))
newdata
```

```{r}
test = sample(nrow(fires_lm), floor(nrow(fires_lm)*.2))
linear_fit = lm(log_area~., data = newdata[-test,])
summary(linear_fit)


# getting MSE for the test
pred = predict(linear_fit, newdata[test,])
mean((pred - fires_lm[test, "area"])**2)
```

# Lasso
```{r}
grid=10^seq(10,-2,length=100)
lasso_mod = glmnet(x = as.matrix(newdata[-test, -"log_area"]), y = as.matrix(newdata[-test, log_area]), alpha=1, lambda = grid)
plot(lasso_mod)

# cross validation
cv_lasso = cv.glmnet(x = as.matrix(newdata[-test, -"log_area"]), y = as.matrix(newdata[-test, log_area]), alpha=1)
plot(cv_lasso)

bestlam = cv_lasso$lambda.min
lasso_pred=predict(lasso_mod,s=bestlam, newx=as.matrix(newdata[-test, -"log_area"]))
mean((lasso_pred-as.matrix(newdata[-test, log_area]))^2 - as.matrix(newdata[-test, log_area]))

out=glmnet(as.matrix(newdata[, -"log_area"]), as.matrix(newdata[, log_area]), alpha=1,lambda=grid)
lasso_coef=predict(out,type="coefficients",s=bestlam)
lasso_coef
```

























